
---
title: "Interpolating irregular data"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 8
vignette: >
  %\VignetteIndexEntry{Interpolating irregular data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is the tour: eight ways to turn scattered measurements into a full grid,
on one set of points, with one grid, so the pictures can be put side by side.

Three other articles go deeper on one thing each. `vignette("grids")` is what a
grid is and how to hand one to another package. `vignette("triangulation")` is
the Delaunay and Voronoi tessellations and the arithmetic behind them.
`vignette("projection")` is what changes when the coordinates change, and how
this all lines up with GDAL.

# Setup

```{r}
library(guerrilla)
library(palr)
```

Everything else is loaded where it is used, so it is visible which method needs
which package.

Read the zooplankton data. 

```{r}
library(readxl)
bw <- read_excel(system.file("extdata", "BW-Zooplankton_env.xls", package= "guerrilla", mustWork = TRUE))
summary(bw[,1:10])
lonlat <- as.matrix(bw[, c("Lon", "Lat")])
val <- bw$temp
```

The values are sea surface temperatures, so the plots use a palette meant for
them, from \pkg{palr}:

```{r}
cols <- sst_pal(24)
```

Plot the temperature data. 

```{r}
plot(val)
```

Create a grid to interpolate onto. The same grid is used for every method
below, so the pictures can be compared.

```{r}
r0 <- grid_spec(lonlat, crs = "EPSG:4326")
r0
```

A grid here is a list of four things and nothing else, and cell centres come
from `grid_xy()`. That is all the interpolation code below knows about grids;
`vignette("grids")` is the whole story if you want it.

```{r}
str(unclass(r0))
head(grid_xy(r0), 3)
```

# Binning

The simplest thing possible is to define a grid and drop the points into it.

```{r}
bingrid <- grid_bin(lonlat, val, r0)
plot(bingrid, col = cols)
points(lonlat, pch = 16, cex = 0.3)
```

Nothing is estimated between the points: a cell with no point in it stays `NA`,
which is most of them. The whole of `grid_bin()` is `vaster::cell_from_xy()`
and `tapply()`, and it is worth having because of the question it forces, which
every other method on this page answers without telling you.

## What happens when two points land in one cell

`fun` is that answer, written down.

```{r}
counts <- grid_bin(lonlat, val, r0, fun = length)
plot(counts)
range(counts$values, na.rm = TRUE)
```

The transect was sampled along a line, so cells on the line hold several
measurements and cells off it hold none. `fun = mean` averages them,
`fun = length` counts them, `fun = function(x) x[1]` keeps whichever came first
in the file. The last one is what a lot of software does by default.

That count grid is the most useful picture in this vignette. Every surface
below fills the whole extent, and every one of them is inventing the parts of
it where this grid is `NA`.

```{r}
plot(grid_bin(lonlat, val, grid_spec(lonlat, dimension = c(15, 12))))
points(lonlat, pch = 16, cex = 0.3)
```

Coarser cells fill more of the grid and say less about any of it. There is no
resolution that fixes this, which is the reason to interpolate.

# Thin plate splines

The surface that passes near the data while bending as little as possible. This
is the engine behind GDAL's `-tps` warping, so it is also what happens when you
georeference an image from ground control points.

```{r}
tpsgrid <- grid_tps(lonlat, val, r0)
plot(tpsgrid, col = cols)
points(lonlat, pch = 16, cex = 0.3)
```

It filled the whole grid, including the corners, where there is no data within
several degrees. That is not a bug and it is not a secret either, because the
same fit will say how sure it is:

```{r}
plot(grid_tps(lonlat, val, r0, statistic = "se"))
points(lonlat, pch = 16, cex = 0.3)
```

The two pictures belong together. The first says what the method thinks and the
second says where it is guessing, and the second one looks a great deal like the
count grid from the previous section.

Read the warning rather than ignoring it. `fields` searched for a smoothing
parameter, found the minimum at the end of its range, and reports 47.5
effective degrees of freedom for 50 data points. That is a spline passing
through the data rather than smoothing it: with fifty points along one transect
there is nothing to average over, so cross validation says do not average.
`method = "REML"` picks the smoothing parameter a different way and is worth
comparing.

```{r}
plot(grid_tps(lonlat, val, r0, method = "REML"), col = cols)
```

`lon.lat` is the one place in this package where what the coordinates mean
changes the arithmetic: a spline bends in the plane of its coordinates, and a
degree of longitude is not a degree of latitude anywhere but the equator. It is
not read off the `crs`, because the guess would be wrong exactly when it
mattered.

```{r}
lltps <- grid_tps(lonlat, val, r0, lon.lat = TRUE)
plot(lltps, col = cols)
```

# Bilinear triangulation

Triangulate the points, and inside each triangle put the plane through its three
corner values. This is what MATLAB calls `griddata(method = "linear")` and GDAL
calls `gdal_grid -a linear`.

```{r}
trigrid <- grid_barycentric(lonlat, val, r0)
plot(trigrid)
points(lonlat, pch = 16, cex = 0.3)
```

It passes exactly through the data and invents nothing beyond it, so cells
outside the convex hull stay `NA`. That is the honest end of the range: no
answer where there is no basis for one.

The arithmetic is barycentric coordinates, which are three numbers that are
both the point-in-triangle test and the interpolation rule at once:

```{r}
bary_weights(cbind(c(0, 1, 0), c(0, 0, 1)),
             rbind(c(0.25, 0.25), c(1/3, 1/3), c(1, 1)))
```

The third point is outside the triangle, which shows as a negative weight rather
than as a separate test. `vignette("triangulation")` goes through this properly,
along with the Delaunay and Voronoi tessellations, the readable R engine, and
what the old `facets()` was really computing.

# Nearest neighbour

```{r}
vor <- grid_voronoi(lonlat, val, r0)
plot(vor)
```

A Voronoi tile is everywhere closer to its own point than to any other, so
filling each tile with its point's value is nearest neighbour drawn rather than
computed. Note that it filled the whole grid where triangulation left the
corners `NA`. Tiles cover the plane, so nothing is ever outside them, and a cell
far from any data still gets a confident answer.

# Inverse distance weighting

A weighted average of the input values, weighted by one over distance to a
power. There is no model here, only a rule.

```{r}
op <- par(mfrow = c(2, 1), mar = c(2, 2, 2, 1))
plot(grid_idw(lonlat, val, r0, idp = 0.5), main = "idp = 0.5")
plot(grid_idw(lonlat, val, r0, idp = 8), main = "idp = 8")
par(op)
```

`idp` is the whole method. Raise it and the nearest point dominates, so the
surface tends towards `grid_voronoi()`; lower it and everything flattens
towards the overall mean. Neither end is more correct than the other and
nothing in the data tells you where to sit between them, which is why the
default of 2 is a convention rather than an answer.

The consequence is that this method cannot report a standard error. It never
claimed to be estimating anything, so there is nothing to be uncertain with.

# Kriging

The same idea taken seriously. The variogram is a picture of how much two
values differ as a function of how far apart they are; kriging fits a curve to
that picture and uses the curve to choose the weights.

```{r}
krigrid <- grid_kriging(lonlat, val, r0)
plot(krigrid, col = cols)
```

Because the weights come from a fitted model, there is a variance to go with
them:

```{r}
plot(grid_kriging(lonlat, val, r0, statistic = "se"))
points(lonlat, pch = 16, cex = 0.3)
```

That surface is usually more informative than the prediction, and it is the
reason to reach for this method rather than IDW.

It is also the only method here that can refuse. Give it values with no spatial
structure and there is no curve to fit, the fit comes back with a negative
range, and it stops:

```{r, error = TRUE}
set.seed(9)
noise <- cbind(runif(120), runif(120))
grid_kriging(noise, rnorm(120))
```

Every other method on this page will interpolate pure noise without complaint
and hand you a picture of it.

# Generalized additive models

```{r}
plot(grid_gam(lonlat, val, r0), col = cols)
```

The default formula is `value ~ s(x, y)`, one isotropic two dimensional smooth.
`mgcv`'s default basis for a two dimensional `s()` is a thin plate regression
spline, so this is the thin plate spline from earlier with fewer basis
functions and the smoothing parameter chosen by REML. On this data the two are
hard to tell apart, which is the point.

The wrong model is instructive too:

```{r}
op <- par(mfrow = c(2, 1), mar = c(2, 2, 2, 1))
plot(grid_gam(lonlat, val, r0), main = "s(x, y)")
plot(grid_gam(lonlat, val, r0, formula = value ~ s(x) + s(y)),
     main = "s(x) + s(y)")
par(op)
```

Additive in x and y means every column of the grid has the same shape as every
other column, up to a shift. It cannot represent a feature that sits in one
place, so it smears anything local into a cross.

# Kernel smoothing

Bin first, estimate second.

```{r}
plot(grid_smooth(lonlat, val, r0, theta = 1), col = cols)
```

Everything else here works from the points; this one goes through a grid on the
way. The binning throws away where inside its cell each point was, and no
amount of smoothing afterwards gets that back. `theta` is the kernel bandwidth,
in the units of the coordinates, and it does all the work.

# Akima (via the interp package)

`interp(method = "linear")` triangulates and interpolates linearly within each
triangle, which is the same thing `grid_barycentric()` does. Worth checking that
claim rather than believing it: on this data the two agree to within 1e-15.

```{r}
library(interp)
akifun <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- grid_spec(xy)
  dm <- grid$dimension; ex <- grid$extent
  ## interp() wants both output axes increasing, and grid rows run top-down.
  ## Hand it a descending yo and it returns all NA, without complaint.
  aklin <- interp(xy[,1], xy[,2], value,
                  vaster::x_from_col(dm, ex, seq_len(dm[1])),
                  rev(vaster::y_from_row(dm, ex, seq_len(dm[2]))), ...)
  grid$values <- as.vector(aklin$z[, rev(seq_len(ncol(aklin$z)))])
  grid
}

akigrid <- akifun(lonlat, val, grid = r0)
plot(akigrid)

## the same estimator as grid_barycentric(), so it had better agree
ok <- !is.na(akigrid$values) & !is.na(trigrid$values)
max(abs(akigrid$values[ok] - trigrid$values[ok]))
```

# What the pictures have in common

Eight methods, one grid, one set of points.

```{r, fig.height = 10}
op <- par(mfrow = c(4, 2), mar = c(2, 2, 2, 1))
plot(bingrid, main = "grid_bin")
plot(vor, main = "grid_voronoi")
plot(trigrid, main = "grid_barycentric")
plot(akigrid, main = "interp::interp")
plot(tpsgrid, main = "grid_tps")
plot(krigrid, main = "grid_kriging")
plot(grid_idw(lonlat, val, r0), main = "grid_idw")
plot(grid_gam(lonlat, val, r0), main = "grid_gam")
par(op)
```

They disagree most where there is no data, and that is the only place the
choice between them matters much. Near the transect they all say roughly what
the measurements say, because there is nothing else for them to say.

Two of them fill only the convex hull and leave the rest `NA`. Four fill the
whole extent from a fitted model, and three of those will tell you how sure
they are if asked. One fills the whole extent from a rule and cannot.

There is no method here that is right. There is a method that makes its
assumptions visible and a method that does not, and the difference is worth
more than the surfaces.

# Where to next

* `vignette("grids")` -- what a grid is, and the converters to raster, terra
  and gdalraster.
* `vignette("triangulation")` -- barycentric weights spelled out, the two
  tessellations drawn, and the readable engine that checks the fast one.
* `vignette("projection")` -- the same data in two coordinate systems, GDAL's
  own gridder, and thin plate splines as georeferencing.
