
---
title: "Interpolating irregular data"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Irregular data interpolation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

Define utility functions for repeated usage, and load required packages. 
```{r}
scl <- function(x) (x - min(x, na.rm = TRUE))/diff(range(x, na.rm = TRUE))
library(guerrilla)
library(raster)
library(gstat)
library(fields)
library(akima)
library(palr)
```

Read the zooplankton data. 

```{r}
library(readxl)
bw <- read_excel(system.file("extdata", "BW-Zooplankton_env.xls", package= "guerrilla", mustWork = TRUE))
summary(bw[,1:10])
lonlat <- as.matrix(bw[, c("Lon", "Lat")])
val <- bw$temp
minmax <- c(-1.8, 40)

rval <- pretty(val, n = 15)

cols <- sstPal(length(rval) - 1)


```

Plot the temperature data. 

```{r}
plot(val)
```

Create a grid to interpolate onto, this can be used repeatedly for many methods. 

```{r}
library(raster)

r0 <- defaultgrid(lonlat)

```



# Simplistic rasterization

The most simple thing possible is to define a grid and simply populate that with values from points. Extend the grid by a small amount. 

```{r}
rasgrid <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- defaultgrid(xy)
   rasterize(xy, grid, field = value, ...)
}
lastgrid <- rasgrid(lonlat, val, r0)

plot(lastgrid)
```

the main impact on the result here is the effect of pixel size, and the behaviour when there is more than one value per pixel. It really depends on the data **compare brokewest with saz?**



# Thin plate spline interpolation. 

```{r}
library(fields)
tpsfun <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- defaultgrid(xy)
  model <- Tps(xy, value, lon.lat = isLonLat(grid), method = "REML")
  interpolate(grid, model, ...)
}
tpsgrid  <- tpsfun(lonlat, val)
tpsgrid <- clamp(tpsgrid, minmax[1], minmax[2])

## this shows the error
tpsSEgrid <- tpsfun(lonlat, val, r0, fun = predictSE)

plot(tpsgrid)
```



# Bilinear triangulation. 

This is identical to Matlab's griddata(, method = "bilinear")

```{r}
library(geometry)
library(sp)
library(guerrilla)


trigrid <- tri_fun(lonlat, val, r0)

plot(trigrid)
```


# Delaunay and Diriclet


What about Delaunay and Dirichlet without interpolation? 


```{r}
## input must be ppp
library(spatstat)
library(maptools)
rasterToOwin <- function(x) {
  owin(c(xmin(x), xmax(x)), c(ymin(x), ymax(x)))
}
colfun <- viridis::viridis
rbase <- r0
d <- bw
xname <- "Lon"
yname <- "Lat"
varname <- "temp"
pObj <- ppp(d[[xname]], d[[yname]], window = rasterToOwin(rbase), marks = d[[varname]])

lltes<-delaunay(pObj)
meanval <- numeric(length(lltes[[3]]))
  for (i in seq_along(meanval)) {
  meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
       }
plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1])
points(d[,c(xname, yname)])

lltes<-dirichlet(pObj)
meanval <- numeric(length(lltes[[3]]))
  for (i in seq_along(meanval)) {
  meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
  }

plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1])
points(d[,c(xname, yname)])

```





Grid interpolation from Dirichlet or Delaunay. This is much slower than the triangulate and cart2bary method above (**Mike needs to illustrate this**) 

```{r}
 
 fcs.dir <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "dirichlet")
 plot(rasterFromXYZ(as.data.frame(fcs.dir)))
 fcs.del <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "delaunay")
 plot(rasterFromXYZ(as.data.frame(fcs.del)))
```

# IDW

```{r}
library(gstat)
idwfun <- function(xy, value, grid = NULL, ...) {
   if (is.null(grid)) grid <- defaultgrid(xy)
  d <- data.frame(lon = xy[,1], lat = xy[,2], value = value)
  coordinates(d) <- ~lon+lat
  proj4string(d) <- proj4string(grid)
  ## inverse distance weighted (IDW)
  mg <- gstat(id = "value", formula = value~1, locations = c("lon", "lat"), data = d, ...) 
  interpolate(grid, mg)          
}


dw <- idwfun(lonlat, val, r0)
dw <- idwfun(lonlat, val, r0, set = list(idp = 1.5))
plot(dw);contour(trigrid, add = TRUE)


```

# Kriging

```{r}

krifun <- function(xy, value, grid = NULL, ...) {
   if (is.null(grid)) grid <- defaultgrid(xy)
  d <- data.frame(lon = xy[,1], lat = xy[,2], value = value)
  coordinates(d) <- ~lon+lat
  proj4string(d) <- proj4string(grid)
 v <- variogram(value~1, d)
 m <- fit.variogram(v, vgm(1, "Sph", 150, 1))
 gOK <- gstat(NULL, "value", value~1, d, model=m)
 interpolate(grid, gOK)

}
krigrid <- krifun(lonlat, val, grid = r0)
plot(krigrid)

```

# fields

```{r}
# out <- as.image(d[,varname], x = d[,c(xname, yname)], nx = 25, ny = 25)
# plot(raster(out))
# look<- image.smooth( out, theta= 2.5) 
# plot(raster(look))

fsmfun <- function(xy, value, grid = NULL) {
    if (is.null(grid)) grid <- defaultgrid(xy)
  nrow(grid) <- ncol(grid) ## apparently required
  out <- as.image(value, x = xy, nx = nrow(grid), ny = ncol(grid))
  raster(image.smooth( out, theta= 2.5) )
}

fsmgrid <- fsmfun(lonlat, val, r0)
plot(fsmgrid)

```

# Akima

```{r}
library(akima)
akifun <- function(xy, value, grid = NULL, ...) {
     if (is.null(grid)) grid <- defaultgrid(xy)
  x0 <- xFromCol(grid)
  y0 <- yFromRow(grid)

  aklin <- interp(xy[,1], xy[,2], value, 
              x0, y0, ...)
  raster(list(x = aklin$x, y = rev(aklin$y), z = aklin$z[,ncol(aklin$z):1]))
}

akigrid <- akifun(lonlat, val, grid = r0)
plot(akigrid)
```


# Hexagons? 

(See hexbin, spatstat). 

## Along-track distance
Previously we used Latitude as our x-axis, this is not very sensible since the track visits the same latitudes at different longitudes. As a simple improvement let's build an along-track axis. 

```{r,eval=FALSE}
# lonname <- "Longitude [degrees_east]"
# latname <- "Latitude [degrees_north]"
# tname <- "yyyy-mm-ddThh:mm:ss.sss"
# d2 <- szt[asub, c(xname, yname, lonname, latname, varname, tname)]
# d2$date <- as.POSIXct(d2[[tname]])
# d2 <- d2[order(d2$date, d2$Depth), ]
# library(geosphere)
# dists <- numeric(nrow(d2)-1)
# for (i in seq_along(dists)) {
#   dists[i] <- spDistsN1(as.matrix(d2[i, c(lonname, latname)]), as.matrix(d2[i+1, c(lonname, latname), drop = FALSE]), longlat = TRUE)
#   }
# d2$dist <- cumsum(c(0, dists))
```


So now. 

```{r}
# xname <- "dist"
# d <- d2
# ex <- extent(range(d[, xname]), range(d[, yname])) * 1.2
# rbase <- raster(ex, ncols = 25, nrows = 25)
# rlast <- rasterize(d[,c(xname, yname)], rbase, field = d[,varname])
# 
# rmean <- rasterize(d[,c(xname, yname)], rbase, field = d[,varname], fun = mean)
# 
# rval <- pretty(d[,varname], n = 15)
# 
# plot(rlast, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
#  plot(rmean, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
# 
# 
# p <- Tps(d[, c(xname, yname)], d[,varname])
# rp <- interpolate(rbase, p)
# rpse <- interpolate(rbase, p, fun = predictSE)
#  plot(rp, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
#  plot(rpse, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
# 
# 
# ## input must be ppp
# library(spatstat)
# library(maptools)
# rasterToOwin <- function(x) {
#   owin(c(xmin(x), xmax(x)), c(ymin(x), ymax(x)))
# }
# pObj <- ppp(d[,xname], d[,yname], window = rasterToOwin(rbase), marks = d[, varname])
# 
# lltes<-delaunay(pObj)
# meanval <- numeric(length(lltes[[3]]))
#   for (i in seq_along(meanval)) {
#   meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
#        }
# plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1])
# points(d[,c(xname, yname)])
# 
# 
# lltes<-dirichlet(pObj)
# meanval <- numeric(length(lltes[[3]]))
#   for (i in seq_along(meanval)) {
#   meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
#        }
# plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1])
# points(d[,c(xname, yname)])
# 
# 
# source("maybetin.R")
# 
# fcs.dir <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "dirichlet")
# plot(rasterFromXYZ(as.data.frame(fcs.dir)))
# fcs.del <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "delaunay")
# plot(rasterFromXYZ(as.data.frame(fcs.del)))
# 
# library(gstat)
# 
# dsp <- d
# coordinates(dsp) <- c(xname, yname)
# ## inverse distance weighted (IDW)
# mg <- gstat(id = "Al", formula = Al~1, locations = c(xname, yname), data=dsp, 
#             nmax=7, set=list(idp = .5))
# z <- interpolate(rbase, mg)
# plot(z)
# 
# ## with log
# v <- variogram(log(Al)~1, dsp)
# m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
# gOK <- gstat(NULL, "logAl", log(Al)~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# ## without log
# v <- variogram(Al~1, dsp)
# m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
# gOK <- gstat(NULL, "Al", Al~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# 
# ## Gaussian
# v <- variogram(Al~1, dsp)
# m <- fit.variogram(v, vgm(1, "Gau", 300, 1))
# gOK <- gstat(NULL, "Al", Al~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# out <- as.image(d[,varname], x = d[,c(xname, yname)], nx = 25, ny = 25)
# plot(raster(out))
# look<- image.smooth( out, theta= 2.5) 
# plot(raster(look))
# 
# 
# library(akima)
# 
# x0 <- xFromCol(rbase)
# y0 <- yFromRow(rbase)
# 
# aklin <- interp(d[,xname], d[,yname], d[,varname], 
#               x0, y0)
# # aksplin <-interp(d[,xname], d[,yname], d[,varname], 
# #               x0, y0, linear = FALSE, extrap = TRUE) 
# rlin <- raster(list(x = aklin$x, y = rev(aklin$y), z = t(aklin$z[,ncol(aklin$z):1])))
# ##rsplin <- raster(aksplin)
# plot(rlin, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
#points(d[, c(xname, yname)], cex = 0.2)
```

