
---
title: "Interpolating irregular data"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 8
vignette: >
  %\VignetteIndexEntry{Interpolating irregular data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

Define utility functions for repeated usage, and load required packages. 

```{r}
library(guerrilla)
library(raster)
library(gstat)
library(fields)
library(interp)
library(palr)
```

Read the zooplankton data. 

```{r}
library(readxl)
bw <- read_excel(system.file("extdata", "BW-Zooplankton_env.xls", package= "guerrilla", mustWork = TRUE))
summary(bw[,1:10])
lonlat <- as.matrix(bw[, c("Lon", "Lat")])
val <- bw$temp
minmax <- c(-1.8, 40)

rval <- pretty(val, n = 15)

cols <- sst_pal(length(rval) - 1)


```

Plot the temperature data. 

```{r}
plot(val)
```

Create a grid to interpolate onto. The same grid is used for every method
below, so the pictures can be compared.

```{r}
r0 <- grid_spec(lonlat, crs = "EPSG:4326")
r0
```

A grid in this package is a list, and that is the whole of it:

```{r}
str(unclass(r0))
```

`dimension` is how many cells across and down, `extent` is the region covered,
`crs` is what the coordinates mean -- here they really are longitude and
latitude, so we said so. There is no fourth secret thing. Every raster class in
R is a container for these same few numbers plus a vector of values, which is
why a plain list is enough to interpolate onto and why converting to any of
those classes is a few lines:

```{r}
as_raster(r0)
```

Cell centres come from `grid_xy()`, and that single function is all the
interpolation code below needs to know about grids:

```{r}
head(grid_xy(r0), 3)
```



# Simplistic rasterization

The most simple thing possible is to define a grid and populate that with values from points. 



Note that `defaultgrid()` covers the exact extent of the input points, so the
outermost samples land on the boundary of the grid. Padding the extent a little
is generally good practice; do it explicitly when you want it.

```{r}
library(raster)
rasgrid <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- grid_spec(xy)
  ## raster::rasterize() wants a raster, so convert at the boundary
  rasterize(xy, as_raster(grid), field = value, ...)
}
lastgrid <- rasgrid(lonlat, val, r0)

plot(lastgrid)
```

The main influence on the result here is purely pixel size, and we must consider the behaviour used when there is more than one value per pixel.  

# Thin plate spline interpolation. 

This is the powerhouse behind GDAL's raster warping, and that suggests that we
could leverage GDAL by creating a data set in the right format for it to warp
to.


```{r}
library(fields)
tpsfun <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- grid_spec(xy)
  r <- as_raster(grid)
  model <- Tps(xy, value, lon.lat = isLonLat(r), method = "REML")
  interpolate(r, model, ...)
}
tpsgrid  <- tpsfun(lonlat, val)
tpsgrid <- clamp(tpsgrid, minmax[1], minmax[2])

## this shows the error
tpsSEgrid <- tpsfun(lonlat, val, r0, fun = predictSE)

plot(tpsgrid, asp = "")
```



# Bilinear triangulation. 

This is identical to Matlab's griddata(, method = "bilinear")

```{r}
trigrid <- tri_fun(lonlat, val, r0)
trigrid

## should be identical, generalizing for mesh3d use
t2 <- mesh_raster(cbind(lonlat, val), r0)
stopifnot(identical(trigrid$values, t2$values))

plot(trigrid)
```


# Delaunay and Diriclet


What about Delaunay and Dirichlet without interpolation? 


```{r}
graphics.off()
## input must be ppp
library(spatstat.geom)

gridToOwin <- function(x) {
  owin(x$extent[1:2], x$extent[3:4])
}
colfun <- viridis::viridis
rbase <- r0
d <- bw
xname <- "Lon"
yname <- "Lat"
varname <- "temp"
pObj <- ppp(d[[xname]], d[[yname]], window = gridToOwin(rbase), marks = d[[varname]])

lltes<-delaunay(pObj)
meanval <- numeric(length(lltes[[3]]))
  for (i in seq_along(meanval)) {
  meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
       }
plot(lltes, do.col = TRUE, values = meanval,
     col = colourmap(colfun(20), range = range(meanval)), main = "")
points(d[,c(xname, yname)])

lltes<-dirichlet(pObj)
meanval <- numeric(length(lltes[[3]]))
  for (i in seq_along(meanval)) {
  meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
  }

plot(lltes, do.col = TRUE, values = meanval,
     col = colourmap(colfun(20), range = range(meanval)), main = "")
points(d[,c(xname, yname)])

```





Grid interpolation from Dirichlet or Delaunay. This is much slower than the triangulate and cart2bary method above (**Mike needs to illustrate this**) 

```{r}
 
 nx <- rbase$dimension[1]; ny <- rbase$dimension[2]
 fcs.dir <- facets(pObj, nx = nx, ny = ny, method = "dirichlet")
 plot(rasterFromXYZ(as.data.frame(fcs.dir)))
 fcs.del <- facets(pObj, nx = nx, ny = ny, method = "delaunay")
 plot(rasterFromXYZ(as.data.frame(fcs.del)))
```

# IDW

```{r}
library(gstat)
idwfun <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- grid_spec(xy)
  r <- as_raster(grid)
  d <- data.frame(lon = xy[,1], lat = xy[,2], value = value)
  coordinates(d) <- ~lon+lat
  proj4string(d) <- proj4string(r)
  ## inverse distance weighted (IDW)
  mg <- gstat(id = "value", formula = value~1, locations = c("lon", "lat"), data = d, ...) 
  interpolate(r, mg)          
}


dw <- idwfun(lonlat, val, r0)
dw <- idwfun(lonlat, val, r0, set = list(idp = 1.5))
plot(dw); contour(as_raster(trigrid), add = TRUE)


```

# Kriging

```{r}

krifun <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- grid_spec(xy)
  r <- as_raster(grid)
  d <- data.frame(lon = xy[,1], lat = xy[,2], value = value)
  coordinates(d) <- ~lon+lat
  proj4string(d) <- proj4string(r)
 v <- variogram(value~1, d)
 m <- fit.variogram(v, vgm(1, "Sph", 150, 1))
 gOK <- gstat(NULL, "value", value~1, d, model=m)
 interpolate(r, gOK)

}
krigrid <- krifun(lonlat, val, grid = r0)
plot(krigrid)

```

# GAM

This needs work, mgcv has more promise than this shows ;)



```{r}
library(mgcv)
gamfit <- gam(temp ~ s(Lon) + s(Lat), data = bw, family = gaussian, method = "REML")
g <- grid_spec(lonlat, dimension = c(50, 50))
gxy <- grid_xy(g)
g$values <- c(predict(gamfit, data.frame(Lon = gxy[, 1], Lat = gxy[, 2])))
plot(g)
```

# fields

```{r}
# out <- as.image(d[,varname], x = d[,c(xname, yname)], nx = 25, ny = 25)
# plot(raster(out))
# look<- image.smooth( out, theta= 2.5) 
# plot(raster(look))

fsmfun <- function(xy, value, grid = NULL) {
  if (is.null(grid)) grid <- grid_spec(xy)
  ## as.image() wants a square grid
  n <- max(grid$dimension)
  out <- as.image(value, x = xy, nx = n, ny = n)
  sm <- image.smooth(out, theta = 2.5)
  ## image.smooth() returns x, y, z with both axes increasing and z[x, y];
  ## a guerrilla grid runs top-down, so flip back
  g <- grid_spec(dimension = c(n, n), extent = grid$extent, crs = grid$crs)
  g$values <- as.vector(sm$z[, rev(seq_len(ncol(sm$z)))])
  g
}

fsmgrid <- fsmfun(lonlat, val, r0)
plot(fsmgrid)

```

# Akima (via the interp package)

`interp(method = "linear")` triangulates and interpolates linearly within each
triangle, which is the same thing `tri_fun()` does. Worth checking that claim
rather than believing it: on this data the two agree to within 1e-15.

```{r}
library(interp)
akifun <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- grid_spec(xy)
  dm <- grid$dimension; ex <- grid$extent
  ## interp() wants both output axes increasing, and grid rows run top-down.
  ## Hand it a descending yo and it returns all NA, without complaint.
  aklin <- interp(xy[,1], xy[,2], value,
                  vaster::x_from_col(dm, ex, seq_len(dm[1])),
                  rev(vaster::y_from_row(dm, ex, seq_len(dm[2]))), ...)
  grid$values <- as.vector(aklin$z[, rev(seq_len(ncol(aklin$z)))])
  grid
}

akigrid <- akifun(lonlat, val, grid = r0)
plot(akigrid)

## the same estimator as tri_fun(), so it had better agree
ok <- !is.na(akigrid$values) & !is.na(trigrid$values)
max(abs(akigrid$values[ok] - trigrid$values[ok]))
```

# Others


Hexagons: see hexbin, spatstat. 


sm package



# Along-track distance
Previously we used Latitude as our x-axis, this is not very sensible since the track visits the same latitudes at different longitudes. As a simple improvement let's build an along-track axis. 

```{r,eval=FALSE}
# lonname <- "Longitude [degrees_east]"
# latname <- "Latitude [degrees_north]"
# tname <- "yyyy-mm-ddThh:mm:ss.sss"
# d2 <- szt[asub, c(xname, yname, lonname, latname, varname, tname)]
# d2$date <- as.POSIXct(d2[[tname]])
# d2 <- d2[order(d2$date, d2$Depth), ]
# library(geosphere)
# dists <- numeric(nrow(d2)-1)
# for (i in seq_along(dists)) {
#   dists[i] <- spDistsN1(as.matrix(d2[i, c(lonname, latname)]), as.matrix(d2[i+1, c(lonname, latname), drop = FALSE]), longlat = TRUE)
#   }
# d2$dist <- cumsum(c(0, dists))
```


So now. 

```{r}
# xname <- "dist"
# d <- d2
# ex <- extent(range(d[, xname]), range(d[, yname])) * 1.2
# rbase <- raster(ex, ncols = 25, nrows = 25)
# rlast <- rasterize(d[,c(xname, yname)], rbase, field = d[,varname])
# 
# rmean <- rasterize(d[,c(xname, yname)], rbase, field = d[,varname], fun = mean)
# 
# rval <- pretty(d[,varname], n = 15)
# 
# plot(rlast, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
#  plot(rmean, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
# 
# 
# p <- Tps(d[, c(xname, yname)], d[,varname])
# rp <- interpolate(rbase, p)
# rpse <- interpolate(rbase, p, fun = predictSE)
#  plot(rp, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
#  plot(rpse, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
# 
# 
# ## input must be ppp
# library(spatstat)

# rasterToOwin <- function(x) {
#   owin(c(xmin(x), xmax(x)), c(ymin(x), ymax(x)))
# }
# pObj <- ppp(d[,xname], d[,yname], window = rasterToOwin(rbase), marks = d[, varname])
# 
# lltes<-delaunay(pObj)
# meanval <- numeric(length(lltes[[3]]))
#   for (i in seq_along(meanval)) {
#   meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
#        }
# plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1])
# points(d[,c(xname, yname)])
# 
# 
# lltes<-dirichlet(pObj)
# meanval <- numeric(length(lltes[[3]]))
#   for (i in seq_along(meanval)) {
#   meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
#        }
# plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1])
# points(d[,c(xname, yname)])
# 
# 
# source("maybetin.R")
# 
# fcs.dir <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "dirichlet")
# plot(rasterFromXYZ(as.data.frame(fcs.dir)))
# fcs.del <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "delaunay")
# plot(rasterFromXYZ(as.data.frame(fcs.del)))
# 
# library(gstat)
# 
# dsp <- d
# coordinates(dsp) <- c(xname, yname)
# ## inverse distance weighted (IDW)
# mg <- gstat(id = "Al", formula = Al~1, locations = c(xname, yname), data=dsp, 
#             nmax=7, set=list(idp = .5))
# z <- interpolate(rbase, mg)
# plot(z)
# 
# ## with log
# v <- variogram(log(Al)~1, dsp)
# m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
# gOK <- gstat(NULL, "logAl", log(Al)~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# ## without log
# v <- variogram(Al~1, dsp)
# m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
# gOK <- gstat(NULL, "Al", Al~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# 
# ## Gaussian
# v <- variogram(Al~1, dsp)
# m <- fit.variogram(v, vgm(1, "Gau", 300, 1))
# gOK <- gstat(NULL, "Al", Al~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# out <- as.image(d[,varname], x = d[,c(xname, yname)], nx = 25, ny = 25)
# plot(raster(out))
# look<- image.smooth( out, theta= 2.5) 
# plot(raster(look))
# 
# 
# library(akima)
# 
# x0 <- xFromCol(rbase)
# y0 <- yFromRow(rbase)
# 
# aklin <- interp(d[,xname], d[,yname], d[,varname], 
#               x0, y0)
# # aksplin <-interp(d[,xname], d[,yname], d[,varname], 
# #               x0, y0, linear = FALSE, extrap = TRUE) 
# rlin <- raster(list(x = aklin$x, y = rev(aklin$y), z = t(aklin$z[,ncol(aklin$z):1])))
# ##rsplin <- raster(aksplin)
# plot(rlin, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
#points(d[, c(xname, yname)], cex = 0.2)
```

