
---
title: "Interpolating irregular data"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 8
vignette: >
  %\VignetteIndexEntry{Interpolating irregular data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

Define utility functions for repeated usage, and load required packages. 

```{r}
scl <- function(x) (x - min(x, na.rm = TRUE))/diff(range(x, na.rm = TRUE))
library(guerrilla)
library(raster)
library(gstat)
library(fields)
library(akima)
library(palr)
```

Read the zooplankton data. 

```{r}
library(readxl)
bw <- read_excel(system.file("extdata", "BW-Zooplankton_env.xls", package= "guerrilla", mustWork = TRUE))
summary(bw[,1:10])
lonlat <- as.matrix(bw[, c("Lon", "Lat")])
val <- bw$temp
minmax <- c(-1.8, 40)

rval <- pretty(val, n = 15)

cols <- sst_pal(length(rval) - 1)


```

Plot the temperature data. 

```{r}
plot(val)
```

Create a grid to interpolate onto, this can be used repeatedly for many methods. 

```{r}
library(raster)

r0 <- defaultgrid(lonlat)

```



# Simplistic rasterization

The most simple thing possible is to define a grid and populate that with values from points. 



We extend the grid by a small amount, which is generally good practice. 

```{r}
rasgrid <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- defaultgrid(xy)
   rasterize(xy, grid, field = value, ...)
}
lastgrid <- rasgrid(lonlat, val, r0)

plot(lastgrid, asp = "")
```

The main influence on the result here is purely pixel size, and we must consider the behaviour used when there is more than one value per pixel.  

# Thin plate spline interpolation. 

This is the powerhouse behind GDAL's raster warping, and that suggests that we
could leverage GDAL by creating a data set in the right format for it to warp
to.


```{r}
library(fields)
tpsfun <- function(xy, value, grid = NULL, ...) {
  if (is.null(grid)) grid <- defaultgrid(xy)
  model <- Tps(xy, value, lon.lat = isLonLat(grid), method = "REML")
  interpolate(grid, model, ...)
}
tpsgrid  <- tpsfun(lonlat, val)
tpsgrid <- clamp(tpsgrid, minmax[1], minmax[2])

## this shows the error
tpsSEgrid <- tpsfun(lonlat, val, r0, fun = predictSE)

plot(tpsgrid, asp = "")
```



# Bilinear triangulation. 

This is identical to Matlab's griddata(, method = "bilinear")

```{r}
library(geometry)
library(sp)
library(guerrilla)


trigrid <- tri_fun(lonlat, val, r0)
## should be identical, generalizing for mesh3d use
t2 <- mesh_raster(cbind(lonlat, val), r0)
plot(trigrid, asp = "")
```


# Delaunay and Diriclet


What about Delaunay and Dirichlet without interpolation? 


```{r}
graphics.off()
## input must be ppp
library(spatstat.geom)
library(maptools)
rasterToOwin <- function(x) {
  owin(c(xmin(x), xmax(x)), c(ymin(x), ymax(x)))
}
colfun <- viridis::viridis
rbase <- r0
d <- bw
xname <- "Lon"
yname <- "Lat"
varname <- "temp"
pObj <- ppp(d[[xname]], d[[yname]], window = rasterToOwin(rbase), marks = d[[varname]])

lltes<-delaunay(pObj)
meanval <- numeric(length(lltes[[3]]))
  for (i in seq_along(meanval)) {
  meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
       }
plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1], asp = "")
points(d[,c(xname, yname)])

lltes<-dirichlet(pObj)
meanval <- numeric(length(lltes[[3]]))
  for (i in seq_along(meanval)) {
  meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
  }

plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1], asp = NA)
points(d[,c(xname, yname)])

```





Grid interpolation from Dirichlet or Delaunay. This is much slower than the triangulate and cart2bary method above (**Mike needs to illustrate this**) 

```{r}
 
 fcs.dir <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "dirichlet")
 plot(rasterFromXYZ(as.data.frame(fcs.dir)), asp = NA)
 fcs.del <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "delaunay")
 plot(rasterFromXYZ(as.data.frame(fcs.del)), asp = NA)
```

# IDW

```{r}
library(gstat)
idwfun <- function(xy, value, grid = NULL, ...) {
   if (is.null(grid)) grid <- defaultgrid(xy)
  d <- data.frame(lon = xy[,1], lat = xy[,2], value = value)
  coordinates(d) <- ~lon+lat
  proj4string(d) <- proj4string(grid)
  ## inverse distance weighted (IDW)
  mg <- gstat(id = "value", formula = value~1, locations = c("lon", "lat"), data = d, ...) 
  interpolate(grid, mg)          
}


dw <- idwfun(lonlat, val, r0)
dw <- idwfun(lonlat, val, r0, set = list(idp = 1.5))
plot(dw, asp = NA);contour(trigrid, add = TRUE)


```

# Kriging

```{r}

krifun <- function(xy, value, grid = NULL, ...) {
   if (is.null(grid)) grid <- defaultgrid(xy)
  d <- data.frame(lon = xy[,1], lat = xy[,2], value = value)
  coordinates(d) <- ~lon+lat
  proj4string(d) <- proj4string(grid)
 v <- variogram(value~1, d)
 m <- fit.variogram(v, vgm(1, "Sph", 150, 1))
 gOK <- gstat(NULL, "value", value~1, d, model=m)
 interpolate(grid, gOK)

}
krigrid <- krifun(lonlat, val, grid = r0)
plot(krigrid, asp = "")

```

# GAM

This needs work, mgcv has more promise than this shows ;)



```{r}
library(mgcv)
library(dplyr)
gamfit <- gam(temp ~ s(Lon)+ s(Lat), data = bw,family = gaussian, method = "REML")
library(raster)
g <- raster(extent(as.matrix(bw[c("Lon", "Lat")])), nrows = 50, ncols = 50)
pred <- predict(gamfit, as.data.frame(g, xy = TRUE) %>% dplyr::rename(Lon = x, Lat = y) %>% dplyr::select(Lon, Lat))
plot(setValues(g, c(pred)), asp = "")
```

# fields

```{r}
# out <- as.image(d[,varname], x = d[,c(xname, yname)], nx = 25, ny = 25)
# plot(raster(out))
# look<- image.smooth( out, theta= 2.5) 
# plot(raster(look))

fsmfun <- function(xy, value, grid = NULL) {
    if (is.null(grid)) grid <- defaultgrid(xy)
  nrow(grid) <- ncol(grid) ## apparently required
  out <- as.image(value, x = xy, nx = nrow(grid), ny = ncol(grid))
  raster(image.smooth( out, theta= 2.5) )
}

fsmgrid <- fsmfun(lonlat, val, r0)
plot(fsmgrid, asp = "")

```

# Akima

```{r}
library(akima)
akifun <- function(xy, value, grid = NULL, ...) {
     if (is.null(grid)) grid <- defaultgrid(xy)
  x0 <- xFromCol(grid)
  y0 <- yFromRow(grid)

  aklin <- interp(xy[,1], xy[,2], value, 
              x0, y0, ...)
  raster(list(x = aklin$x, y = rev(aklin$y), z = aklin$z[,ncol(aklin$z):1]))
}

akigrid <- akifun(lonlat, val, grid = r0)
plot(akigrid, asp = "")
```

# Others


Hexagons: see hexbin, spatstat. 


sm package



# Along-track distance
Previously we used Latitude as our x-axis, this is not very sensible since the track visits the same latitudes at different longitudes. As a simple improvement let's build an along-track axis. 

```{r,eval=FALSE}
# lonname <- "Longitude [degrees_east]"
# latname <- "Latitude [degrees_north]"
# tname <- "yyyy-mm-ddThh:mm:ss.sss"
# d2 <- szt[asub, c(xname, yname, lonname, latname, varname, tname)]
# d2$date <- as.POSIXct(d2[[tname]])
# d2 <- d2[order(d2$date, d2$Depth), ]
# library(geosphere)
# dists <- numeric(nrow(d2)-1)
# for (i in seq_along(dists)) {
#   dists[i] <- spDistsN1(as.matrix(d2[i, c(lonname, latname)]), as.matrix(d2[i+1, c(lonname, latname), drop = FALSE]), longlat = TRUE)
#   }
# d2$dist <- cumsum(c(0, dists))
```


So now. 

```{r}
# xname <- "dist"
# d <- d2
# ex <- extent(range(d[, xname]), range(d[, yname])) * 1.2
# rbase <- raster(ex, ncols = 25, nrows = 25)
# rlast <- rasterize(d[,c(xname, yname)], rbase, field = d[,varname])
# 
# rmean <- rasterize(d[,c(xname, yname)], rbase, field = d[,varname], fun = mean)
# 
# rval <- pretty(d[,varname], n = 15)
# 
# plot(rlast, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
#  plot(rmean, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
# 
# 
# p <- Tps(d[, c(xname, yname)], d[,varname])
# rp <- interpolate(rbase, p)
# rpse <- interpolate(rbase, p, fun = predictSE)
#  plot(rp, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
#  plot(rpse, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
# points(d[, c(xname, yname)], cex = 0.2)
# 
# 
# ## input must be ppp
# library(spatstat)
# library(maptools)
# rasterToOwin <- function(x) {
#   owin(c(xmin(x), xmax(x)), c(ymin(x), ymax(x)))
# }
# pObj <- ppp(d[,xname], d[,yname], window = rasterToOwin(rbase), marks = d[, varname])
# 
# lltes<-delaunay(pObj)
# meanval <- numeric(length(lltes[[3]]))
#   for (i in seq_along(meanval)) {
#   meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
#        }
# plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1])
# points(d[,c(xname, yname)])
# 
# 
# lltes<-dirichlet(pObj)
# meanval <- numeric(length(lltes[[3]]))
#   for (i in seq_along(meanval)) {
#   meanval[i] <- mean(marks(pObj[lltes[[3]][[i]]]))
#        }
# plot(as(lltes, "SpatialPolygons"), col = colfun(20)[scl(meanval)*19 + 1])
# points(d[,c(xname, yname)])
# 
# 
# source("maybetin.R")
# 
# fcs.dir <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "dirichlet")
# plot(rasterFromXYZ(as.data.frame(fcs.dir)))
# fcs.del <- facets(pObj, nx = ncol(rbase) , ny = nrow(rbase), method = "delaunay")
# plot(rasterFromXYZ(as.data.frame(fcs.del)))
# 
# library(gstat)
# 
# dsp <- d
# coordinates(dsp) <- c(xname, yname)
# ## inverse distance weighted (IDW)
# mg <- gstat(id = "Al", formula = Al~1, locations = c(xname, yname), data=dsp, 
#             nmax=7, set=list(idp = .5))
# z <- interpolate(rbase, mg)
# plot(z)
# 
# ## with log
# v <- variogram(log(Al)~1, dsp)
# m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
# gOK <- gstat(NULL, "logAl", log(Al)~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# ## without log
# v <- variogram(Al~1, dsp)
# m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
# gOK <- gstat(NULL, "Al", Al~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# 
# ## Gaussian
# v <- variogram(Al~1, dsp)
# m <- fit.variogram(v, vgm(1, "Gau", 300, 1))
# gOK <- gstat(NULL, "Al", Al~1, dsp, model=m)
# OK <- interpolate(rbase, gOK)
# plot(OK)
# 
# out <- as.image(d[,varname], x = d[,c(xname, yname)], nx = 25, ny = 25)
# plot(raster(out))
# look<- image.smooth( out, theta= 2.5) 
# plot(raster(look))
# 
# 
# library(akima)
# 
# x0 <- xFromCol(rbase)
# y0 <- yFromRow(rbase)
# 
# aklin <- interp(d[,xname], d[,yname], d[,varname], 
#               x0, y0)
# # aksplin <-interp(d[,xname], d[,yname], d[,varname], 
# #               x0, y0, linear = FALSE, extrap = TRUE) 
# rlin <- raster(list(x = aklin$x, y = rev(aklin$y), z = t(aklin$z[,ncol(aklin$z):1])))
# ##rsplin <- raster(aksplin)
# plot(rlin, col = colfun(length(rval) - 1), breaks = rval, asp = 1)
#points(d[, c(xname, yname)], cex = 0.2)
```

