---
title: "Coordinates, projections, and GDAL"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Coordinates, projections, and GDAL}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Nothing in this package's interpolation looks at a coordinate reference system.
The arithmetic is planar: triangles, distances, weights, all computed on the
numbers you hand it. A `crs` carried in on the input is carried out on the
result, untouched, because it is a label on the answer rather than an input to
it.

That is a deliberate choice and it is not the same as saying projection does not
matter. It matters a great deal. It just is not the interpolation's business,
and this article is about what it *is* the business of.

```{r}
library(guerrilla)
library(reproj)
library(readxl)

bw <- read_excel(system.file("extdata", "BW-Zooplankton_env.xls",
                             package = "guerrilla", mustWork = TRUE))
lonlat <- as.matrix(bw[, c("Lon", "Lat")])
val <- bw$temp
```

# The same points, in two coordinate systems

The transect runs about 50 degrees of longitude at 65 degrees south, where a
degree of longitude is under half the length of a degree of latitude. Project
it to a local equal-area and the shape changes:

```{r}
mid <- c(mean(range(lonlat[, 1])), mean(range(lonlat[, 2])))
laea <- sprintf("+proj=laea +lon_0=%f +lat_0=%f +datum=WGS84", mid[1], mid[2])
xy <- reproj_xy(lonlat, laea, source = "EPSG:4326")

op <- par(mfrow = c(2, 1), mar = c(2, 3, 2, 1))
plot(lonlat, pch = 16, cex = 0.4, asp = 1, main = "longitude and latitude")
plot(xy, pch = 16, cex = 0.4, asp = 1, main = "local equal area, metres")
par(op)
```

Interpolate on each, and the surfaces are not the same surface.

```{r}
gl <- grid_spec(lonlat, dimension = c(120, 100), crs = "EPSG:4326")
gp <- grid_spec(xy, dimension = c(120, 100), crs = laea)

a <- grid_barycentric(lonlat, val, gl)
b <- grid_barycentric(xy, val, gp)

op <- par(mfrow = c(2, 1), mar = c(2, 3, 2, 1))
plot(a, main = "interpolated in degrees")
plot(b, main = "interpolated in metres")
par(op)
```

To compare them properly, take the cell centres of the first grid, project them,
and read the second surface there:

```{r}
sampled <- rep(NA_real_, grid_ncell(gl))
at <- reproj_xy(grid_xy(gl), laea, source = "EPSG:4326")
cell <- vaster::cell_from_xy(gp$dimension, gp$extent, at)
ok <- !is.na(cell)
sampled[ok] <- b$values[cell[ok]]

both <- !is.na(sampled) & !is.na(a$values)
c(compared = sum(both),
  max_difference = max(abs(sampled[both] - a$values[both])),
  rms_difference = sqrt(mean((sampled[both] - a$values[both])^2)),
  data_range = diff(range(val)))
```

The two agree at the data points, because both are exact there. In between they
differ by up to a fifth of the whole range of the data, from nothing but the
choice of coordinates.

They also cover different areas:

```{r}
c(only_in_degrees = sum(!is.na(a$values) & is.na(sampled)),
  only_in_metres = sum(is.na(a$values) & !is.na(sampled)))
```

Triangulation is built out of straight lines and a convex hull, and a straight
line in one projection is not a straight line in another. So the triangles are
different triangles, the hull is a different hull, and the region the answer
covers changes as well as the answer.

None of that is something the interpolation could have decided for you. Which
projection is right depends on what the values mean and what you intend to do
with the surface, and only you know that. What this package can do is not
pretend the question was settled.

# GDAL does this too

GDAL's `gdal_grid` interpolates scattered data, and it has been doing it since
long before any of this. Four of its algorithms have a twin here.

```{r}
g <- grid_spec(lonlat, dimension = c(60, 50), crs = "EPSG:4326")
gdal_linear <- grid_gdal(lonlat, val, g, "linear:radius=0.0")
ours        <- grid_barycentric(lonlat, val, g)

identical(is.na(gdal_linear$values), is.na(ours$values))
max(abs(gdal_linear$values - ours$values), na.rm = TRUE)
```

Agreement to floating point, between GDAL's C++ and this package's call into
Qhull. And for nearest neighbour, agreement exactly:

```{r}
identical(grid_gdal(lonlat, val, g, "nearest")$values,
          grid_voronoi(lonlat, val, g)$values)
```

That one is worth a second look, because the two arrive by different routes.
GDAL searches for the closest point; `grid_voronoi()` builds the Voronoi
tessellation with GEOS and asks which tile each cell falls in. Same answer, to
the last bit, three thousand times.

Inverse distance is the same story with a little more room in it:

```{r}
max(abs(grid_gdal(lonlat, val, g, "invdist:power=2.0:smoothing=0.0")$values -
        grid_idw(lonlat, val, g)$values))
```

Four independent implementations of linear interpolation over a Delaunay
triangulation now agree on this data: this package's `tsearch()` path, its
readable R path, `interp::interp()`, and GDAL. That is worth more than any one
of them being convincing on its own.

## Two GDAL defaults to know about

`gdal_grid`'s `linear` algorithm does not stop at the convex hull unless you
tell it to. The default `radius` is -1, an infinite search, so a cell in no
triangle silently takes the value of the nearest point:

```{r}
c(default = sum(is.na(grid_gdal(lonlat, val, g)$values)),
  radius_zero = sum(is.na(grid_gdal(lonlat, val, g, "linear:radius=0.0")$values)))
```

And cells it cannot estimate are filled with 0, with nothing recorded to say
that 0 is not a measurement. `grid_gdal()` sets `nodata=nan` for you unless you
set one yourself, which is why the numbers above come back as `NA`.

## Where gdal_grid lives

`gdal_grid` is reachable from R only through `sf::gdal_utils()`.
\pkg{gdalraster} wraps `warp`, `translate` and `rasterize` but not `GDALGrid`,
and the unified `gdal` command line introduced in GDAL 3.11 has no grid
subcommand for it to wrap. So `grid_gdal()` needs \pkg{sf}, and that is the
only reason it does.

# Thin plate splines, in the other direction

`grid_tps()` fits a spline to `(x, y) -> value`. GDAL's `-tps` warping fits a
spline to `(pixel, line) -> (x, y)`. Same estimator, applied to coordinates
instead of measurements: georeferencing an image from ground control points is
interpolation, and it is the interpolation in this package.

The claim is checkable. Build an image whose two bands hold its own column and
row index, so that after warping we can read off which input pixel every output
cell came from:

```{r, eval = FALSE}
library(gdalraster)
nx <- 120L; ny <- 100L
src <- tempfile(fileext = ".tif")
create(format = "GTiff", dst_filename = src, xsize = nx, ysize = ny,
       nbands = 2L, dataType = "Float64")
ds <- new(GDALRaster, src, read_only = FALSE)
ds$write(band = 1L, xoff = 0L, yoff = 0L, xsize = nx, ysize = ny,
         rasterData = as.numeric(rep(seq_len(nx), times = ny)))
ds$write(band = 2L, xoff = 0L, yoff = 0L, xsize = nx, ysize = ny,
         rasterData = as.numeric(rep(seq_len(ny), each = nx)))
ds$close()
```

Give it sixteen ground control points, mapping a lattice of pixels through the
same longlat to equal-area transform used above. That mapping is smooth and it
is not affine, which is exactly the case `-tps` exists for.

```{r, eval = FALSE}
gp <- as.matrix(expand.grid(px = seq(1, nx, length.out = 4),
                            py = seq(1, ny, length.out = 4)))
lon <- 30 + (gp[, "px"] - 1) / (nx - 1) * 50
lat <- -69 + (gp[, "py"] - 1) / (ny - 1) * 8
mapxy <- reproj_xy(cbind(lon, lat), laea, source = "EPSG:4326")

## GDAL counts pixel and line from the top left CORNER of the image, so the
## centre of the first pixel is at 0.5, not at 1. Our column index is 1-based
## and names the pixel itself. That half pixel is the oldest bug in
## georeferencing and it is worth being deliberate about.
gcp <- as.vector(rbind("-gcp",
                       format(gp[, "px"] - 0.5, digits = 15),
                       format(gp[, "py"] - 0.5, digits = 15),
                       format(mapxy[, 1], digits = 15),
                       format(mapxy[, 2], digits = 15)))

withgcp <- tempfile(fileext = ".tif")
warped <- tempfile(fileext = ".tif")
translate(src, withgcp, cl_arg = c(gcp, "-a_srs", laea), quiet = TRUE)
warp(withgcp, warped, t_srs = laea, cl_arg = c("-tps", "-r", "near"),
     quiet = TRUE)
```

Now fit the same thing with \pkg{fields}, from the same sixteen points, and
compare where GDAL and R each think a given map position came from.

```{r, eval = FALSE}
library(fields)
fx <- Tps(mapxy, gp[, "px"], lambda = 0)   ## lambda = 0 interpolates exactly,
fy <- Tps(mapxy, gp[, "py"], lambda = 0)   ## which is what GDAL's tps does

out <- new(GDALRaster, warped)
ox <- out$getRasterXSize(); oy <- out$getRasterYSize()
gt <- out$getGeoTransform()
col_from <- out$read(band = 1L, xoff = 0L, yoff = 0L, xsize = ox, ysize = oy,
                     out_xsize = ox, out_ysize = oy)
row_from <- out$read(band = 2L, xoff = 0L, yoff = 0L, xsize = ox, ysize = oy,
                     out_xsize = ox, out_ysize = oy)
out$close()

oc <- rep(seq_len(ox), times = oy) - 0.5
orw <- rep(seq_len(oy), each = ox) - 0.5
mx <- gt[1] + oc * gt[2] + orw * gt[3]
my <- gt[4] + oc * gt[5] + orw * gt[6]

filled <- is.finite(col_from) & col_from > 0
dcol <- predict(fx, cbind(mx[filled], my[filled])) - col_from[filled]
drow <- predict(fy, cbind(mx[filled], my[filled])) - row_from[filled]
c(cells = sum(filled),
  mean_column_offset = mean(dcol), sd_column = sd(dcol),
  mean_row_offset = mean(drow), sd_row = sd(drow))
```

```
#>              cells mean_column_offset          sd_column
#>       8631.000000           0.001064           0.383321
#>   mean_row_offset             sd_row
#>          0.026572           0.431023
```

A mean offset of one thousandth of a pixel, over eight thousand cells. The
remaining scatter of about four tenths of a pixel is the nearest-neighbour
resampling: the warped image records a whole pixel index, and the spline
predicts a fractional one. It does not shrink if you make the output finer,
which is how you know it is the rounding and not the fit.

So GDAL's `-tps` and `fields::Tps()` are the same estimator. `grid_tps()` and
image georeferencing are the same operation pointed at different quantities,
and if you understand one you understand the other.

The chunks above are not run when this vignette is built, because they write
files. Run them yourself; they work.

# A note on coordinate magnitude

One thing did have to change to make projected coordinates work at all.
`geometry::tsearch()` builds a quadtree over the points, and on some coordinate
ranges the insertion fails outright:

```{r, error = TRUE}
big <- cbind((lonlat[, 1] - mid[1]) * 1000, (lonlat[, 2] - mid[2]) * 1000)
tri <- geometry::delaunayn(big)
at <- grid_xy(grid_spec(big, dimension = c(50, 40)))
geometry::tsearch(big[, 1], big[, 2], tri, at[, 1], at[, 2], bary = TRUE)
```

The same points scaled by 10000 instead of 1000 are fine, so it is not a
threshold to steer around. The fix is to remove the variable: barycentric
weights do not change when a triangle and the point inside it are moved and
scaled together, so `grid_barycentric()` centres and scales the coordinates
before the search and the answer is unaffected.

```{r}
head(grid_barycentric(big, val, grid_spec(big, dimension = c(50, 40)))$values, 3)
```

This is the kind of thing that only shows up when you leave degrees behind,
which is a reason to try your work in a projection even when you did not need
one.
