---
title: "Triangles, tiles, and the weights between them"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Triangles, tiles, and the weights between them}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Two of the methods in this package come from tessellating the plane rather than
fitting anything: cut the space up around the data points, then let each piece
answer for the region it covers. They give the two simplest honest answers
there are, and the arithmetic behind both fits on one page.

```{r}
library(guerrilla)
library(readxl)
bw <- read_excel(system.file("extdata", "BW-Zooplankton_env.xls",
                             package = "guerrilla", mustWork = TRUE))
lonlat <- as.matrix(bw[, c("Lon", "Lat")])
val <- bw$temp
g <- grid_spec(lonlat, crs = "EPSG:4326")
```

# The two tessellations

GEOS builds both, and they are duals of each other.

```{r}
library(geos)
pts <- as_geos_geometry(wk::xy(lonlat[, 1], lonlat[, 2]))
coll <- geos_make_collection(pts)

op <- par(mfrow = c(2, 1), mar = c(2, 2, 2, 1))
plot(geos_unnest(geos_delaunay_triangles(coll), keep_multi = FALSE),
     col = NA, border = "grey40", main = "Delaunay triangles")
points(lonlat, pch = 16, cex = 0.4)
plot(geos_unnest(geos_voronoi_polygons(coll), keep_multi = FALSE),
     col = NA, border = "grey40", main = "Voronoi tiles")
points(lonlat, pch = 16, cex = 0.4)
par(op)
```

They answer different questions. A Delaunay triangle has three data points at
its corners, so a value inside it can be a blend of three. A Voronoi tile has
one data point in it and consists of everywhere closer to that point than to any
other, so a value inside it can only be that one point's value.

Which means the Voronoi tessellation *is* nearest neighbour:

```{r}
plot(grid_voronoi(lonlat, val, g))
points(lonlat, pch = 16, cex = 0.3)
```

# Barycentric coordinates

The Delaunay case needs one idea. For a point inside a triangle, the
barycentric weights say how much of each corner the point is made of. They sum
to one, and they are all non-negative exactly when the point is inside -- so the
same three numbers are both a point-in-triangle test and the interpolation rule.

```{r}
tri <- cbind(c(0, 1, 0), c(0, 0, 1))
bary_weights(tri, rbind(c(0.25, 0.25), c(1/3, 1/3), c(1, 1)))
```

The first point is a quarter of the way in from two corners; the second is the
centroid, made equally of all three; the third is outside, which shows up as a
negative weight rather than as a separate test.

The whole of `bary_weights()` is Cramer's rule on a two by two system, six
lines of arithmetic. Reading it is a better use of ten minutes than reading any
description of it, including this one.

Interpolating is then the weighted sum of the corner values:

```{r}
tri_grid <- grid_barycentric(lonlat, val, g)
plot(tri_grid)
```

Inside each triangle the surface is the plane through its three corner values,
so the result is continuous, passes exactly through the data, and invents
nothing beyond it. Cells outside the convex hull are in no triangle and stay
`NA` -- unlike the Voronoi fill above, which covers everything. That is
convenient, and it is also nearest neighbour's main way of lying to you: a cell
far from any data still gets a confident answer.

# Two engines, one answer

`grid_barycentric()` normally gets its result from one call to
`geometry::tsearch()`, which finds the containing triangle for every cell and
its weights in a single pass of C. That is the right way to compute it and the
wrong way to learn it, so the same job is also written out in R:
`find_triangle()` to locate each cell, `bary_weights()` to weight it.

```{r}
slow <- grid_barycentric(lonlat, val, g, engine = "R")
ok <- !is.na(slow$values)
max(abs(slow$values[ok] - tri_grid$values[ok]))
```

The readable one is much slower and agrees to floating point. It is not there
for speed, it is there so the claim "this is what the C is doing" can be
checked rather than believed.

`find_triangle()` is worth a look on its own, because the search is the part
people assume is hard. It asks GEOS which triangles a point could be in, using
an STRtree index, then uses `bary_weights()` to decide which one it actually is.

```{r}
triangles <- geometry::delaunayn(lonlat)
find_triangle(lonlat, triangles, rbind(c(50, -65), c(0, 0)))
```

The second point is nowhere near the data, and gets `NA`.

# The long way round, and a bug it revealed

Three points determine a plane, and least squares through exactly three points
is that plane. So fitting `value ~ x + y` inside each Delaunay triangle must
give the same surface as barycentric interpolation:

```{r}
lm_grid <- grid_facet_lm(lonlat, val, g)
both <- !is.na(lm_grid$values) & !is.na(tri_grid$values)
max(abs(lm_grid$values[both] - tri_grid$values[both]))
```

It does, and it is enormously slower, and seeing that happen once is worth the
wait.

This package used to offer that as `facets()`, with a Dirichlet variant
alongside the Delaunay one. The Dirichlet variant was not doing what it looked
like it was doing. A Voronoi tile holds exactly one point, so the per-tile
`lm(value ~ x + y)` had one observation and three parameters. It fitted an
intercept, predicted that single value across the whole tile, and produced --
bit for bit -- nearest neighbour, computed by running a linear model per tile in
an R loop.

Brute force nearest neighbour, one line, gives exactly the same numbers:

```{r}
at <- grid_xy(g)
nearest <- apply(at, 1, function(p) which.min((lonlat[, 1] - p[1])^2 +
                                              (lonlat[, 2] - p[2])^2))
identical(grid_voronoi(lonlat, val, g)$values, val[nearest])
```

`facets()` is still exported and still works, marked superseded. The lesson is
not that it was slow. It is that a method can be named after its machinery
rather than its answer, and then nobody notices for years what answer it gives.
Every function added in this package is named for what it computes, because of
this one.

# The same thing, from a mesh

A grid is one particular mesh: a regular one, with values at the vertices. So
turning an arbitrary mesh of triangles into a grid is resampling one mesh onto
another, and it is the same barycentric interpolation with the triangulation
step already done.

```{r}
mesh <- structure(list(vb = rbind(t(cbind(lonlat, val)), 1),
                       it = t(geometry::delaunayn(lonlat)),
                       primitivetypes = "triangle"),
                  class = c("mesh3d", "shape3d"))
plot(mesh_raster(mesh, grid = g))
```

Same numbers, because it is the same triangles:

```{r}
identical(mesh_raster(mesh, grid = g)$values, tri_grid$values)
```

`mesh_raster()` runs the argument backwards from everything else here, which is
why it is worth having. Given points it triangulates them; given a mesh it
already has the triangles and only has to interpolate. Anything \pkg{Rvcg} or
\pkg{anglr} can hand you as a `mesh3d` can go straight onto a grid.
