---
title: "Mapping cruise data from BROKE-West, with R"
author: "Michael Sumner, Kerrie Swadling and Tomas Remenyi"
date: "8/27/2015"
output: html_document
---

Here we explore some of the many ways to interpolate from scattered data, in R. 

To simplify the examples we create a function for each specific type of interpolation used, and each function has inputs "xy", "value" and (optionally) "grid". 


The example data we use is transect-based voyage data from the Aurora Australis. 

```{r,echo=FALSE,message=FALSE}
library(fields)
```

Get some bathymetric data and save it (see Appendix), we will use this to complete our map. 

Load the bathymetry data. 
```{r}
load("data/broke_bathy.RData")
```

Read the zooplankton data. 

```{r}
library(readxl)
bw <- read_excel("data/BW-Zooplankton_env.xls")
summary(bw[,1:10])
lonlat <- as.matrix(bw[, c("Lon", "Lat")])
val <- bw$temp
minmax <- c(-1.8, 40)
```

Plot the temperature data. 

```{r}
plot(val)
```

Create a grid to interpolate onto, this can be used repeatedly for many methods. 

```{r}
library(raster)
defaultgrid <- function(xy, ncol = 60, nrow = 50, prj = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") {
  raster(extent(xy), ncol = 60, nrow = 50, crs = prj)
}
r0 <- defaultgrid(lonlat)

```

Thin plate spline interpolation. 

```{r}
library(fields)
tpsfun <- function(xy, value, grid = NULL) {
  if (is.null(grid)) grid <- defaultgrid(xy)
  model <- Tps(xy, value, lon.lat = isLonLat(grid), method = "REML")
  interpolate(grid, model)
}
tpsgrid  <- tpsfun(lonlat, val)
tpsgrid <- clamp(tpsgrid, minmax[1], minmax[2])

```


library(palr)
spal <- palr::sstPal(palette = TRUE)
plot(NA, asp = NA, xlab = "Longitude", ylab = "Latitude", xlim = c(20, 90), ylim = c(-70, -60))
plot(temp, add = TRUE, col = spal$cols, breaks = spal$breaks, legend = FALSE)
contour(temp, add = TRUE)
contour(bathy,  lwd = 2, lev = -1, add = TRUE); ##c(-5000, -4000, -3000, -500, -200))
points(xy, pch = 21, cex = 2, bg = sstPal(bw$temp))

# use model to predict values at all locations
plot(NA, asp = NA, xlab = "Longitude", ylab = "Latitude", xlim = c(20, 90), ylim = c(-70, -60))
plot(ifd, add = TRUE)
contour(ifd, add = TRUE)
contour(bathy,  lwd = 2, lev = -1, add = TRUE); ##c(-5000, -4000, -3000, -500, -200))

# use model to predict values at all locations
plot(NA, asp = NA, xlab = "Longitude", ylab = "Latitude", xlim = c(20, 90), ylim = c(-70, -60))
pal <- palr::chlPal(palette = TRUE)
plot(chl, add = TRUE, col = pal$cols, breaks = pal$breaks, legend = FALSE)
contour(chl, add = TRUE)
contour(bathy,  lwd = 2, lev = -1, add = TRUE); ##c(-5000, -4000, -3000, -500, -200))
points(xy, pch = 21, cex = 2, bg = chlPal(bw$`chl a mg/m2)`/1000))


```

Using triangulation etc. 

```{r}

library(gstat)
bwd <- as.data.frame(bw)
bwd$x <- bwd$Lon
bwd$y <- bwd$Lat
mg <- gstat(id = "temp", formula = temp~1, locations = ~x+y, data=bwd, nmax = 30, set=list(idp = .35))

z <- interpolate(r0, mg)
plot(z)




library(rgdal)
library(RTriangle)
prj <- "+proj=laea +lon_0=54 +lat_0=-63"
pxy <- project(xy, prj)
tri <- RTriangle::triangulate(pslg(P = xy))
#rxy <- project(coordinates(r0), prj)
rxy <- coordinates(r0)
 tri_pip <- function(tri, pts) {
   ps <- lapply(split(tri$T, seq(nrow(tri$T))), function(x) Polygon(tri$P[c(x, x[1]), ]))
   sp <- lapply(seq_along(ps), function(x) Polygons(ps[x], x))
   spp <- SpatialPolygons(sp, proj4string = CRS(proj4string(pts)))
   
   over(pts, spp)
 }
scl <- function(x) (x - min(x))/diff(range(x))
 pid <- tri_pip(tri, SpatialPoints(rxy, proj4string = CRS(prj)))
 ok <- !is.na(pid)
 weighted <- numeric(length(pid))
 library(geometry)
for (i in which(ok)) {
  #print(i)
  tripts <- tri$P[tri$T[pid[i], ], ]
  rpts <- rxy[pid == pid[i] & !is.na(pid) , ]
  wgts <- cart2bary(tripts, rpts)
  vals <- matrix(bw$temp[tri$T[pid[i], ]], ncol = 3, nrow = nrow(wgts), byrow = TRUE)
  weighted[pid == pid[i] & !is.na(pid)] <- rowSums(vals * wgts)
}
  plot(setValues(r0, weighted), asp = NA)
 
  ```
# Appendix


 
```{r,eval=FALSE}
library(raadtools)
bathy1 <- aggregate(readtopo("etopo2", xylim = extent(15, 95, -75, -55)) , fact = 2, fun = mean)
gf <- focalWeight(bathy1, res(bathy1) * 6, "Gauss")
bathy <- focal(bathy1, gf)
save(bathy, file = "data/broke_bathy.RData")

```
```{r,eval=FALSE,echo=FALSE}
library(rmatio)
bathy0 <- read.mat("afternoon_exercise/broke_bathy.mat")
xynames <- c("lon", "lat")
len <- length(bathy0)
## you cannot change the xlim/ylim in R after the fact
lon <- unlist(bathy0[seq(1, len, by = 2)])
lat <- unlist(bathy0[seq(2, len, by = 2)])

```
